\bf Range Analysis for the IonMonkey JavaScript engine

### 2012-04-30 project-writings/paper.tex
Write up some of the equations for the stuff we do.
Write a lot of paper

@@

OTHER OPTIMIZATIONS WE DO

OUR CONTRIBUTIONS

@@

@@

>>>>

\newcommand{\Squash}{\mathsf{Squash}}
\newcommand{\Clamp}{\mathsf{Clamp}}
\newcommand{\Range}{\mathsf{Range}}
\let\x\cdot

@@

OTHER OPTIMIZATIONS WE DO??

The core of our analysis system is essentially an implementation of
Gough and Klaeren's static range analysis algorithm for bounds check
elimination.
We identify our technical contributions as:
\begin{itemize}
\item An implementation of Gough and Klaeren's range analysis in a
  modern industrial-strength SSA-based compiler.
\item A different algorithm for inserting and removing $\beta$ nodes
  that used by Gough and Klaeren that takes advantage of the data
  structures of a modern SSA compiler.
\item To facilitate the above, extended IonMonkey with a fast
  (constant time, one integer per basic block memory overhead) 
  mechanism for determining whether a block is dominated by another.
\item Extending the analysis to produce the information needed for
  overflow check elimination by differentiating between
  ``maximum/minimum 32-bit int values'' and ``infinity'' (possible
  overflow).
\item The observation that when doing range computation with the above
  distinction, the input ranges can be always be ``squashed'' to only
  finite ranges, since if overflow had occured, the code would have
  bailed out.
\item An implementation of overflow check elimination using the above
  analyses.
\end{itemize}

@@

\section{IonMonkey Overview}

@@

\subsection{Ranges and Range Arithmetic}

In Single Static Assignment form, each computation has a distinct
name. We take advantage of this to assign each name (that is, each
computation) a ``range''; the ranges in our system are a contiguous
subset of the set ${-\infty} \cup [-2^{31}, 2^{31}-1] \cup {\infty}$
that represent the possible results a computation may have. A range
that includes $\pm \infty$ indicates that the associated
computation may overflow in that direction.

\begin{figure}[ht]
\begin{eqnarray*}
Z &=& {-\infty} \cup [-2^{31}, 2^{31}-1] \cup {\infty} \\
\Clamp(n) &=& \begin{cases}
-\infty &\text{if } n < -2^{31} \\
\infty &\text{if } n > -2^{31}-1 \\
n&\text{otherwise } \\
\end{cases}\\
\Clamp([l, h]) &=& [\Clamp(l), \Clamp(h)] \\
{[x_l, x_h] \cup [y_l, y_h]} &=& [\min(x_l, y_l), \max(x_h, y_h)] \\
{[x_l, x_h] \cap [y_l, y_h]} &=& [\max(x_l, y_l), \min(x_h, y_h)] \\
{[x_l, x_h] + [y_l, y_h]} &=& \Clamp ([x_l + y_l, x_h + y_h]) \\
{[x_l, x_h] - [y_l, y_h]} &=& \Clamp ([x_l - y_h, x_h + y_l]) \\
{[x_l, x_h] \x [y_l, y_h]} &=& 
\begin{tabular}{rl}
$\Clamp ([$&$\min(x_l \x y_l, x_l \x y_h, x_h \x y_l, x_h \x y_h),$\\
&$\max(x_l \x y_l, x_l \x y_h, x_h \x y_l, x_h \x y_h)])$ \\
\end{tabular}
\end{eqnarray*}

\caption{Range arithmetic definitions}
\label{fig:range_arith}
\end{figure}

We can define arithmetical and set operations on these ranges in the
standard way, ``clamping'' values outside the range of a 32-bit signed
integer to $\pm \infty$. Some definitions for this are are shown in
Figure \ref{fig:range_arith}.

\subsection{Conditional Branches and $\beta$ nobes}

An important source of range information that requires care to take
advantage of is conditional branches. Consider the code in Figure
\ref{fig:branch}. All three arithmetic operations in this code can be
proven to not overflow, but to do this it is not sufficient to simply
associate each name with a range, since the information differs
between basic blocks. An obvious approach is to instead associate
ranges with name/basic block pairs, allowing representation of
flow-sensitive range information. This solution, however, is
unsatisfying. One of the major benefits of SSA-style analysis is
eliminating the need for this sort of tracking; by giving each
computation a unique name, SSA reduces the need for tracking
information about variables across time; adopting this approach would
undo this.

The approach we took, again following \cite{Gough94eliminatingrange},
is to instead attack the problem in ``SSA style'', by introducing
$\beta$-nodes, a new form of pseudo-operation that associates range
information with a value. $\beta$ nodes take one argument and
additionall have an auxillary constant range associated with
them. Operationally, $\beta$-nodes are simply copies, but the
invariant is maintained that the output value will fall inside the
range given by the $\beta$ node. Gough and Klaeren refer to SSA
extended with $\beta$ nodes as XSA.

\begin{figure}[ht]
\begin{center}
\begin{verbatim}
var y;
if (x < 0) {
    y = x + 2000000000;
} else {
    if (x < 1000000000) {
        y = x*2;
    } else {
        y = x - 3000000000;
    }
}
\end{verbatim}
\end{center}
\caption{Code that requires information about branches to properly analyze}
\label{fig:branch}
\end{figure}

\begin{figure}[ht]
\begin{center}
\begin{verbatim}
if (x < 0) {
    x1 = Beta(x, [INT_MIN, -1]);
    y1 = x1 + 2000000000;
} else {
    x2 = Beta(x, [0, INT_MAX]);
    if (x2 < 1000000000) {
        x3 = Beta(x2, [INT_MIN, 999999999]);
        y2 = x3*2;
    } else {
        x4 = Beta(x2, [1000000000, INT_MAX]);
        y3 = x4 - 3000000000;
    }
    y4 = Phi(y2, y3);
}
y = Phi(y1, y4);
\end{verbatim}
\end{center}
\caption{Code from Figure \ref{fig:branch}, in XSA form}
\label{fig:xsa}
\end{figure}

\subsection{Range Analysis}

\begin{figure}[ht]
\begin{eqnarray*}
\Squash(n) &=& \begin{cases}
 -2^{31} &\text{if } n = -\infty \\
2^{31}-1 &\text{if } n = \infty \\
n&\text{otherwise } \\
\end{cases}\\
\Squash([l, h]) &=& [\Squash(l), \Squash(h)]
\end{eqnarray*}
%
\begin{eqnarray*}
x = \Phi(y, z) &\Rightarrow&
    \Range(x) = \Squash(\Range(y)) \cup \Squash(\Range(z)) \\
x = \beta(y, [l, h]) &\Rightarrow& \Range(x) = \Squash(\Range(y)) \cap [l, h] \\
x = y + z &\Rightarrow& \Range(x) = \Squash(\Range(y)) + \Squash(\Range(z)) \\
x = y - z &\Rightarrow& \Range(x) = \Squash(\Range(y)) - \Squash(\Range(z)) \\
x = y \x z &\Rightarrow& \Range(x) = \Squash(\Range(y)) \x \Squash(\Range(z)) \\
x = c &\Rightarrow& \Range(x) = [c, c] \\
\end{eqnarray*}

\caption{Range analysis equations}
\label{fig:range_anal}
\end{figure}
